---
title: "Exploratory Data Analysis on the Wine Quality Dataset"
output: github_document
---


**Prepare the Problem**
```{r}
# 1. Prepare Problem
# a) Load libraries
library(ggplot2)
library(caret)
library(corrplot)
library(gridExtra)
library(e1071)
# b) Load dataset
wine.white <- read.table("winequality-white.csv",header = T,sep=";")
wine.red <- read.table("winequality-red.csv",header = T,sep=";")
wine.white$type <- "White"
wine.red$type <- "Red"
wine.full <- rbind(wine.white,wine.red)
# c) Split-out validation dataset
trainRows <- createDataPartition(1:nrow(wine.full),p=.7,list = F)
trainSet <- wine.full[trainRows,]
testSet <- wine.full[-trainRows,]
```


**Exploratory Data Analysis**

```{r}
#Functions for plotting the data
plotHist <- function(data_in, i) {
  data <- data.frame(x=data_in[[i]])
  p <- ggplot(data=data, aes(x=factor(x))) + stat_count() + xlab(colnames(data_in)[i]) + theme_light() + 
    theme(axis.text.x = element_text(angle = 90, hjust =1))
  return (p)
}

plotDen <- function(data_in, i){
  data <- data.frame(x=data_in[[i]])
  p <- ggplot(data= data) + geom_line(aes(x = x), stat = 'density', size = 1,alpha = 1.0) +
    xlab(paste0((colnames(data_in)[i]), '\n', 'Skewness: ',round(skewness(data_in[[i]], na.rm = TRUE), 2))) + theme_light() 
  return(p)
  
}

plotCorr <- function(data_in, i){
  data <- data.frame(x = data_in[[i]], quality = data_in$quality)
  p <- ggplot(data, aes(x = x, y = quality)) + geom_point(shape = 1, na.rm = TRUE) + geom_smooth(method = lm ) + xlab(paste0(colnames(data_in)[i], '\n', 'R-Squared: ', round(cor(data_in[[i]], data$quality), 2))) + theme_light()
  return(suppressWarnings(p))
}

doPlots <- function(data_in, fun, ii, ncol=3) {
  pp <- list()
  for (i in ii) {
    p <- fun(data_in=data_in, i=i)
    pp <- c(pp, list(p))
  }
  do.call("grid.arrange", c(pp, ncol=ncol))
}
```


**Density Plots for Continuous Variables**
The density plots of the features indicate that the features are skewed.

```{r}
doPlots(trainSet[,1:11],fun=plotDen,ii=1:4,ncol = 2)
```
```{r}
doPlots(trainSet[,1:11],fun=plotDen,ii=5:8,ncol = 2)
```
```{r}
doPlots(trainSet[,1:11],fun=plotDen,ii=9:11,ncol = 2)
```

**Histograms for the Categorical Variables**
```{r}
doPlots(trainSet[,12:13],fun=plotHist,ii=1:2,ncol = 2)
```
The distribution of quality, the response variable, is already normal so there is no more need to transform the data.


**Correlation Matrix**
```{r}
#How are the variables corelated to one another?
correlations <-cor(trainSet[,sapply(trainSet,is.numeric)])
corrplot(correlations,method = "square")
```

The correlation matrix above suggests that none of the predictors are highly correlated to the quality of the wine

Notable variables with relatively high correlation with quality:
* volatile.acidity
* chlorides
* density
* alcohol

**ANOVA Test for Categorical Variables**

```{r}
type.aov <- aov(quality ~ type,data=trainSet)
anova(type.aov)
```
```{r}
TukeyHSD(type.aov)
```

```{r}
plot(TukeyHSD(type.aov))
```
The analysis of variance (ANOVA) on the type categorical variable yielded significant variation as seen above. A post hoc Tukey Test showed that the white wines and red wines differed significantly at p < .05.
